Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2020-01-22 Manuscript/manuscript.Rmd
results edits

## A resolution-specific approach for predicting TAD boundaries

We analyzed publically available replicate Hi-C data sets from the GM12878 and K562 cell lines (Supplementary Table 1). Topologically associating domains (TADs) were subsequently called using the ARROWHEAD pipeline [@durand2016juicer] at five data resolutions: 5 kb 10 kb, 25 kb, 50 kb, and 100 kb. <!--After removing TADs that exceeded 2 md in length (biologically plausible maximum TAD size [@dixon2012topological; @sofueva2013cohesin]),--> The results presented are for the GM12878 cell line, with K562 results presented in the Supplementary material. We obtained a total of 8052, 7676, 4670, 2349, and 1031 called TADs across each resolution respectively (Table 1). From the called TADs we obtained a total of 15468, 14253, 8363, 4224, and 1883 unique TAD boundaries (Table 2). The resolution-specific approach we developed to create TAD vs non-TAD boundary regions amenable for binary classification (See Methods) resulted in a total of 535363, 267682, 107073, 53537, and 26768 genomic bins. TAD boundary regions were defined as regions on the linear genome where a genomic bin contained a called TAD boundary ($Y = 1$), while non-TAD boundary regions were defined as genomic bins that did not contain a TAD boundary ($Y = 0$) (Figure 1).

## TAD vs. non-TAD boundary class imbalance is severe irrespectively of Hi-C data resolution

Under this construction, the class imbalance among the genomic bins was found to be both severe and invariant toward resolution. Specifically, we found the imbalance between TAD boundary and non-TAD boundary regions to be 0.03, 0.05, 0.08, 0.08, and 0.07 for each resolution respectively (Table 2). Furthermore, the data summaries presented in Table 1 were found to be similar to those exibited by the K562 cell line (Supplementary Table 3).

@@

![**Illustrating the resolution-specific data construction and feature engineering used for downstream modeling.** Unique boundaries are denoted by the blue dots. The linear genome was subsequently binned into resolution-specific intervals. The response vector **Y** was defined as a genomic bin that either overlapped with a TAD boundary ($Y = 1$) or did not ($Y = 0$). Three types of associations between bins and genomic annotations (green shapes) were then used to build the feature space of random forest classification models include _Overlap Counts_ (OC), _Overlap Percent_ (OP), and ($log_{2}$) _Distance_. ](Figures/model_construction_and_predictor_types.png)

>>>>

## Developing a computational framework for optimal TAD boundary prediction

We developed a computational framework, refered to as **TADRF**, aimed at determining the optimal set of data level characteristics necessary for TAD boundary prediction using Random Forest classification models. Our framework is resolution-specific and uses called TAD boundaries as ground truth to establish TAD-boundary and non-TAD boundary regions. Under our framework, we binned the linear genome at intervals specific to the Hi-C data resolution used to call TADs. TAD boundary regions were defined as regions on the linear genome where a genomic bin contained a called TAD boundary ($Y = 1$), while non-TAD boundary regions were defined as genomic bins that did not contain a TAD boundary ($Y = 0$) (Figure 1 A). Under this construction, the number of non-TAD boundary regions are exceed the number of TAD boundary regions. To cope with the imbalance among the class of genomic bins, we evaluated three re-sampling techniques (in addition to no re-sampling) before predicting TAD boundaries. **Random over-sampling** (ROS) was defined as sampling with replacement from TAD boundary regions. **Random under-sampling** (RUS) was defined as sampling with replacement from non-TAD boundary regions. Lastly, we considered **Synthetic minority over-sampling technique** (SMOTE), which is a combination of both random over- and under-sampling [@chawla2002smote].

Various one-dimensional cell line-specific ChIP-seq tracks (genomic elements) were used to construct the feature space for the random forests including cis-regulatory elements, histone modifications, and transcription factor binding sites. Four different feature engineering procedures were developed for each genomic element (Figure 1 B). **Peak Signal** was defined as the [???include definition]. Counts of overlaps (**Overlap Counts**, OC) were defined as the total number of genomic annotation regions that overlapped within a given genomic bin. Percents of overlaps (**Overlap Percent**, OP) were defined as the proportion of a bin overlapped by an annotation region. **Distance** was defined as log2 transformed number of base pairs between between the centers of bins and regions defining genomic annotations. A log2-transformation of distance was used to account for the skewness present in the distributions of genomic distances (Supplementary Figure 1). These feature engineering procedures were performed to improve TAD boundary predictions by accounting for direct (OC), proportional (Signal/OP), and spatial (distance) relationships between TAD boundaries and genomic annotations.

![**Illustrating the resolution-specific data construction and feature engineering used for downstream random forest modeling.** (A) The linear genome was subsequently binned into resolution-specific intervals. The response vector **Y** was defined as a genomic bin that either overlapped with a TAD boundary ($Y = 1$) or did not ($Y = 0$). (B) Four types of associations between bins and genomic annotations (green shapes) were then used to build the feature space including _Peak Signal_ (Signal), _Overlap Counts_ (OC), _Overlap Percent_ (OP), and ($log_{2}$) _Distance_. ](Figures/model_construction_and_predictor_types.png)

We applied our framework to publically available replicate Hi-C data sets from the GM12878 and K562 cell lines (Supplementary Table 1). Topologically associating domains (TADs) were subsequently called using the ARROWHEAD pipeline [@durand2016juicer] at five data resolutions: 5 kb 10 kb, 25 kb, 50 kb, and 100 kb. The results presented are for the GM12878 cell line, with K562 results presented in the Supplementary material. We obtained a total of 8052, 7676, 4670, 2349, and 1031 called TADs across each resolution respectively (Table 1). From the called TADs we obtained a total of 15468, 14253, 8363, 4224, and 1883 unique TAD boundaries (Table 2). The resolution-specific approach we developed to create TAD vs non-TAD boundary regions amenable for binary classification (See Methods) resulted in a total of 535363, 267682, 107073, 53537, and 26768 genomic bins. The class imbalance among the genomic bins was found to be both severe and invariant toward resolution. Specifically, we found the imbalance between TAD boundary and non-TAD boundary regions to be 0.03, 0.05, 0.08, 0.08, and 0.07 for each resolution respectively (Table 2). Furthermore, the data summaries presented in Table 1 were found to be similar to those exibited by the K562 cell line (Supplementary Table 3).

@@
